\chapter{Research Objectives}\label{sec:research-objectives}

\section{Main Objective}\label{sec:main-objective}

To systematically evaluate and compare classical and AI-based relay strategies for two-hop cooperative communication, and to determine the optimal relay architecture as a function of channel conditions, SNR regime, and computational constraints.

\section{Research Hypotheses}\label{sec:research-hypotheses}

Based on the theoretical analysis in Chapter~\ref{sec:experiments}, this thesis tests the following hypotheses:

\textbf{H1 (Selective AI advantage at low SNR).} Some AI-based relay strategies achieve lower BER than AF and, on selected channels, lower BER than DF at low SNR (0--4 dB), where the non-linear denoising learned by the neural network may provide an advantage over AF's noise amplification and DF's hard-decision errors.

\emph{Rationale:} At low SNR, the Bayes-optimal relay function \(f^*(y) = \tanh(y/\sigma^2)\) is a smooth non-linear mapping that differs substantially from both AF's linear scaling and DF's hard sign function. A neural network can approximate \(f^*\) closely, while the classical methods cannot.

\textbf{H2 (DF dominance at high SNR).} The DF relay achieves the lowest BER at medium-to-high SNR (\(\geq 6\) dB), outperforming all AI methods.

\emph{Rationale:} At high SNR, \(f^*(y) \approx \text{sign}(y)\), which is exactly the DF operation. AI relays introduce a small but non-zero approximation error, so DF should be optimal in this regime.

\textbf{H3 (Inverted-U complexity curve).} There exists an optimal model size for relay denoising beyond which performance degrades due to overfitting. Specifically, models with $\sim$(100–200 parameters achieve performance comparable to models with 10–100$)\times$ more parameters.

\emph{Rationale:} The bias-variance analysis (Section~\ref{sec:theoretical-basis-universal-approximation-and-denoising}) predicts that the low-complexity target function \(f^*\) requires minimal model capacity. Excess capacity increases variance without reducing bias.

\textbf{H4 (Architecture convergence at equal scale).} When all AI models are normalized to the same parameter count ($\sim$3,000), the performance differences between architectures narrow significantly, indicating that parameter count is a more important factor than architectural choice.

\emph{Rationale:} All architectures are universal approximators and the target function is simple. At equal capacity, architectural inductive biases provide diminishing returns.

\textbf{H5 (SSM speed advantage at long context).} Mamba-2 SSD trains significantly faster than Mamba S6 at longer context lengths (\(n \gg 11\)) due to chunk-parallel computation, while S6 is faster at the short relay window (\(n = 11\)).

\emph{Rationale:} The crossover between sequential S6 (\(O(n)\) serial steps) and chunk-parallel SSD (\(O(n/L)\) parallel matmuls of size \(L\)) depends on the ratio of kernel-launch overhead to arithmetic cost.

\textbf{H6 (Equalization gains are additive to relay gains).} The BER improvement from better equalization (ZF \(\to\) MMSE \(\to\) SIC) and the improvement from better relay processing (AF \(\to\) AI) are approximately additive in the dB domain.

\emph{Rationale:} The relay operates on Hop 1 (denoising) and the equalizer operates on Hop 2 (stream separation). Since these are independent processing stages, their contributions to BER reduction should be approximately independent.

\section{Specific Objectives}\label{sec:specific-objectives}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Implement and compare nine relay strategies} spanning four learning paradigms: no learning (AF, DF), supervised learning (MLP, Hybrid), generative modeling (VAE, CGAN), and sequence modeling (Transformer, Mamba S6, Mamba-2 SSD).
\item
  \textbf{Evaluate across six channel/topology configurations:} AWGN, Rayleigh fading, and Rician fading (K=3) channels in SISO topology, and 2×2 MIMO Rayleigh with ZF, MMSE, and SIC equalization.
\item
  \textbf{Investigate the complexity--performance trade-off} by testing model architectures ranging from 0 parameters (classical) to 26,179 parameters (Mamba-2 SSD), and by conducting a normalized comparison at approximately 3,000 parameters.
\item
  \textbf{Determine whether state space models outperform attention mechanisms} for relay signal processing by comparing Mamba S6 and Transformer architectures at both original and normalized parameter counts, and by benchmarking at extended context lengths.
\item
  \textbf{Identify practical deployment recommendations} for selecting the appropriate relay strategy given specific operational constraints (SNR range, computational budget, channel environment).
\end{enumerate}

\section{Scope and Delimitations}\label{sec:scope-and-delimitations}

The following delimitations define the scope of this study:

\begin{itemize}
\tightlist
\item
  \textbf{Modulation:} Primary experiments use BPSK. Extension experiments with QPSK and 16-QAM are presented in Section~\ref{sec:higher-order-modulation-scalability-constellation-aware-training}, and 16-PSK in Section~\ref{sec:input-normalization-and-csi-injection}, to evaluate hypothesis generalisability; however, 64-QAM and higher-order constellations are deferred to future work.
\item
  \textbf{Channel knowledge:} Perfect CSI is assumed at the receiver. Channel estimation errors are not modeled.
\item
  \textbf{Relay topology:} Single relay, two-hop, half-duplex. Multi-relay and full-duplex configurations are excluded.
\item
  \textbf{MIMO configuration:} \(2 \times 2\) spatial multiplexing with Rayleigh fading. Larger arrays and beamforming are not considered.
\item
  \textbf{Training regime:} Offline training on synthetic data. Online adaptation is not implemented.
\end{itemize}

These delimitations are chosen to enable a clean comparison of relay processing functions in isolation, without confounding factors from protocol-level or system-level complexity.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

