%% ============================================================
%% Chapter 5: Experiments
%% Structure: Master config table + conclusion-only sections
%% All figures and tables are in Appendix (ch09_appendices.tex)
%% ============================================================
\chapter{Experiments}\label{sec:experiments}

This chapter presents the experimental evaluation of nine relay strategies across eight
systematic experiments. Each experiment is described by its goal, configuration, and
conclusion only. All supporting figures and numerical tables are consolidated in
Appendix~\ref{chap:app-experiments} to maintain narrative clarity.

%% ============================================================
%% Master Configuration Table
%% ============================================================
\section{Experiment Overview}\label{sec:experiment-overview}

Table~\ref{tbl:master-config} summarises the configuration space explored across all
eight experiments. Each row defines the topology, modulation, channel, equalizer, and
neural-network settings used; outcomes are reported in the per-experiment sections below
and detailed in Appendix~\ref{chap:app-experiments}.

\begin{table}[H]
\centering
\caption{Master experiment configuration table. All nine relay strategies
(AF, DF, MLP, Hybrid, VAE, CGAN, Transformer, Mamba~S6, Mamba-2~SSD) are evaluated
unless otherwise noted. ``CA'' = Constellation-Aware; ``E2E'' = End-to-End.}
\label{tbl:master-config}
\small
\begin{tabular}{@{}p{0.04\linewidth}p{0.18\linewidth}p{0.10\linewidth}p{0.12\linewidth}p{0.12\linewidth}p{0.10\linewidth}p{0.22\linewidth}@{}}
\toprule
\textbf{E\#} & \textbf{Experiment} & \textbf{Topology} & \textbf{Modulation} & \textbf{Channel} & \textbf{Equalizer} & \textbf{Special Setting} \\
\midrule
E1 & Channel Model Validation & SISO, MIMO 2$\times$2 & BPSK & AWGN, Rayleigh, Rician & None / ZF / MMSE / SIC & Theory vs.\ simulation \\
\midrule
E2 & SISO BPSK Baseline & SISO & BPSK & AWGN, Rayleigh, Rician & None & Activation: \texttt{tanh} \\
\midrule
E3 & MIMO 2$\times$2 BPSK & MIMO 2$\times$2 & BPSK & Rayleigh & ZF, MMSE, SIC & Activation: \texttt{tanh} \\
\midrule
E4 & Parameter Normalisation & SISO, MIMO 2$\times$2 & BPSK & AWGN, Rayleigh, Rician, MIMO & None / ZF / MMSE / SIC & $\sim$3\,K normalised params \\
\midrule
E5 & Higher-Order Modulation & SISO & QPSK, 16-QAM & AWGN, Rayleigh & None & CA activations: \texttt{tanh}, \texttt{hardtanh}, scaled \texttt{tanh}, scaled sigmoid \\
\midrule
E6 & CSI Injection \& LayerNorm & SISO & 16-QAM, 16-PSK & Rayleigh & None & 48 variants: Baseline / CSI / LN / CSI+LN \\
\midrule
E7 & 16-Class 2D Classification & SISO & 16-QAM & AWGN & None & Joint 2D softmax (16-class) \\
\midrule
E8 & End-to-End Optimisation & SISO & Learned ($M{=}16$) & Rayleigh & ZF (receiver) & MLP autoencoder (E2E) \\
\bottomrule
\end{tabular}
\end{table}

%% ============================================================
%% E1: Channel Model Validation
%% ============================================================
\section{E1: Channel Model Validation}\label{sec:channel-model-validation}

\textbf{Goal.}
Validate the simulation framework against closed-form theoretical BER expressions to
ensure baseline accuracy before evaluating AI relays.

\textbf{Configuration.}
Topology: SISO and MIMO 2$\times$2.
Modulation: BPSK.
Channels: AWGN, Rayleigh, Rician ($K{=}3$).
Equalizers (MIMO only): ZF, MMSE, SIC.
Demodulation: hard decision.

\textbf{Conclusion.}
Monte Carlo simulations match theoretical predictions within the 95\,\% confidence
interval at all SNR points for every channel and equalizer combination.
The simulation framework is validated and used as the ground truth for all subsequent
experiments.
Full validation charts are provided in Appendix~\ref{sec:app-e1}.

%% ============================================================
%% E2: SISO BPSK Baseline
%% ============================================================
\section{E2: SISO BPSK Performance (Baseline Relay Comparison)}\label{sec:siso-bpsk-performance-baseline-relay-comparison}

\textbf{Goal.}
Evaluate baseline classical and AI-based relay strategies on single-antenna
configurations across different fading environments.

\textbf{Configuration.}
Topology: SISO.
Modulation: BPSK.
Channels: AWGN, Rayleigh, Rician ($K{=}3$).
Equalizers: none.
NN architectures: MLP, Hybrid, VAE, CGAN, Transformer, Mamba~S6, Mamba-2~SSD.
Activation: \texttt{tanh}.

\textbf{Conclusion.}
AI relays selectively outperform AF and, on AWGN and Rician channels, DF at low SNR
(0--4\,dB).
Under Rayleigh fading, classical DF dominates even at low SNR.
Across all channels, DF remains dominant at medium-to-high SNR ($\geq 6$\,dB),
matching or exceeding all AI methods with zero parameters.
BER curves and numerical tables are in Appendix~\ref{sec:app-e2}
(Figures~\ref{fig:fig9}--\ref{fig:fig11}).

%% ============================================================
%% E3: MIMO 2x2 BPSK
%% ============================================================
\section{E3: MIMO 2$\times$2 BPSK Performance}\label{sec:mimo-2x2-bpsk-performance}

\textbf{Goal.}
Evaluate relay strategies under spatial multiplexing and various interference
cancellation techniques.

\textbf{Configuration.}
Topology: MIMO 2$\times$2.
Modulation: BPSK.
Channel: Rayleigh.
Equalizers: ZF, MMSE, SIC.
NN architectures: all nine.
Activation: \texttt{tanh}.

\textbf{Conclusion.}
The MIMO equalization hierarchy (ZF $<$ MMSE $<$ SIC) holds for all relay types.
The AI advantage at low SNR is preserved under ZF and MMSE (Mamba~S6 achieves the
lowest BER), but under SIC equalization, classical DF provides the lowest BER at all
low-to-medium SNR points.
Relay processing gains and MIMO equalization gains are additive.
BER curves and numerical tables are in Appendix~\ref{sec:app-e3}
(Figures~\ref{fig:fig12}--\ref{fig:fig14}).

%% ============================================================
%% E4: Parameter Normalisation
%% ============================================================
\section{E4: Parameter Normalisation \& Complexity Trade-off}\label{sec:parameter-normalization-complexity-trade-off}

\textbf{Goal.}
Isolate architectural inductive biases from parameter-count effects and characterise
the complexity--performance trade-off for relay denoising.

\textbf{Configuration.}
Topology: SISO and MIMO 2$\times$2.
Modulation: BPSK.
Channels: AWGN, Rayleigh, Rician, MIMO (ZF, MMSE, SIC).
NN architectures: all nine, normalised to $\sim$3\,000 parameters and at original
sizes (169 to 26\,K).
Activation: \texttt{tanh}.

\textbf{Conclusion.}
The relay denoising task exhibits an inverted-U complexity relationship: a minimal
169-parameter MLP matches models 140$\times$ larger, while excessive parameters
($>$11\,K) lead to overfitting.
At a normalised scale of 3\,000 parameters, the performance gap between feedforward
and sequence architectures narrows to $\sim$1\,\% BER, indicating that parameter
count rather than architectural choice is the primary performance driver.
Generative VAE is a consistent underperformer due to probabilistic overhead.
BER curves, complexity charts, and numerical tables are in Appendix~\ref{sec:app-e4}
(Figures~\ref{fig:fig15}--\ref{fig:fig20}).

%% ============================================================
%% E5: Higher-Order Modulation
%% ============================================================
\section{E5: Higher-Order Modulation Scalability (Constellation-Aware Training)}\label{sec:higher-order-modulation-scalability-constellation-aware-training}

\textbf{Goal.}
Evaluate the generalisability of BPSK-trained relays to complex constellations and
resolve the multi-level amplitude bottleneck.

\textbf{Configuration.}
Topology: SISO.
Modulations: QPSK, 16-QAM.
Channels: AWGN, Rayleigh.
Equalizers: none.
NN architectures: all nine.
Activations: \texttt{tanh}, linear, \texttt{hardtanh}, scaled \texttt{tanh}, scaled
sigmoid.
Special case: constellation-aware training.

\textbf{Conclusion.}
QPSK performance mirrors BPSK perfectly due to I/Q independence.
On 16-QAM, standard \texttt{tanh} compression causes a severe, irreducible BER floor
($\approx$0.22 at 16\,dB).
Replacing \texttt{tanh} with constellation-aware bounded activations
(\texttt{hardtanh}, scaled \texttt{tanh}) bounded to the precise signal amplitude
($3/\sqrt{10}$) and retraining eliminates this floor.
Sequence models benefit most, reducing their BER floor by 5$\times$, though a gap to
classical DF persists due to per-axis error accumulation.
BER curves, activation comparisons, and numerical tables are in
Appendix~\ref{sec:app-e5}
(Figures~\ref{fig:fig21}--\ref{fig:fig36}).

%% ============================================================
%% E6: CSI Injection & LayerNorm
%% ============================================================
\section{E6: Input Normalisation and CSI Injection}\label{sec:input-normalization-and-csi-injection}

\textbf{Goal.}
Evaluate the impact of explicit Channel State Information (CSI) injection and Input
Layer Normalisation across 48 distinct neural relay configurations for 16-QAM and
16-PSK.

\textbf{Configuration.}
Topology: SISO.
Modulations: 16-QAM, 16-PSK.
Channel: Rayleigh.
Equalizers: none.
NN architectures: Transformer, Mamba~S6, Mamba-2~SSD.
Variants: Baseline / CSI / LayerNorm / CSI+LayerNorm $\times$ 4 activations
$\times$ 3 architectures $\times$ 2 modulations = 48 configurations.

\textbf{Conclusion.}
A clear dichotomy emerges: explicit CSI injection significantly improves performance
for phase-only modulations (16-PSK) by resolving phase ambiguity, but actively
degrades performance for amplitude-dependent modulations (16-QAM) by disrupting
amplitude-decision boundaries.
Input Layer Normalisation is universally beneficial across all architectures and
modulations.
The full 48-variant sweep, BER charts, and training curves are in
Appendix~\ref{chap:app-csi}
(Figures~\ref{fig:fig39}--\ref{fig:fig45}).

%% ============================================================
%% E7: 16-Class 2D Classification
%% ============================================================
\section{E7: 16-Class 2D Classification for QAM16}\label{sec:class-2d-classification-for-qam16}

\textbf{Goal.}
Eliminate the structural BER floor imposed by per-axis I/Q splitting for 16-QAM by
utilising full 2D decision boundaries.

\textbf{Configuration.}
Topology: SISO.
Modulation: 16-QAM.
Channel: AWGN.
Equalizers: none.
NN architectures: MLP, VAE, Transformer, Mamba~S6, Mamba-2~SSD.
Activation: softmax (implicit via cross-entropy loss).
Special case: joint 16-class 2D classification.

\textbf{Conclusion.}
Treating the relay as a joint 16-point classifier over the full 2D constellation space
completely eliminates the structural 4-class BER floor.
For the first time, neural variants (VAE, Transformer, MLP) matched classical DF
performance at high SNR, achieving near-zero BER at 20\,dB.
This proves the previous BER floor was an artefact of I/Q splitting, not a fundamental
limitation of neural relays.
BER curves, bar charts, and numerical tables are in Appendix~\ref{sec:app-e7}
(Figures~\ref{fig:fig50}--\ref{fig:fig53}).

%% ============================================================
%% E8: End-to-End Optimisation
%% ============================================================
\section{E8: End-to-End Joint Optimisation}\label{sec:end-to-end-joint-optimization}

\textbf{Goal.}
Compare the modular neural relay approach against a fully joint transmitter--receiver
autoencoder.

\textbf{Configuration.}
Topology: SISO.
Modulation: learned latent space ($M{=}16$, power-constrained).
Channel: Rayleigh.
Equalizer: ZF (receiver side).
NN architecture: MLP autoencoder (encoder/decoder).
Special case: end-to-end (E2E) optimisation.

\textbf{Conclusion.}
The E2E autoencoder underperforms both the theoretical limits of classical 16-QAM and
the modular two-hop DF relay across all SNR points (67--141\,\% higher BER).
The network fails to discover a constellation geometry that surpasses the classical
square grid under single-antenna Rayleigh fading, demonstrating that black-box deep
learning is inefficient compared to modular designs that leverage classical signal
processing for modulation and equalization.
BER curves, learned constellation, and training loss are in Appendix~\ref{sec:app-e8}
(Figures~\ref{fig:fig46}--\ref{fig:fig49}).