\chapter {Summary}
\noindent
This work examined classical and neural network–based relay strategies for \textit{two-hop cooperative communication} using a unified simulation framework. The objective was to analyze performance–complexity tradeoffs and determine when learning-based relays provide practical advantages over analytically derived methods.
The entire study is carried out on one canonical setup, fixed in advance and never departed from: a two-hop SISO link over i.i.d.\ Rayleigh fast fading, complex baseband, BPSK, uncoded BER, with the relay function as the only varied axis. The study showed the following:
\begin{enumerate}
    \item Neural relays yield a consistent performance improvement over amplify-and-forward (AF) at low signal-to-noise ratio (SNR), where their learned soft denoising avoids AF's noise amplification (H1).
    \item Classical \textit{decode-and-forward (DF)} matched or outperformed all neural approaches at \textbf{medium-to-high SNR} while requiring no trainable parameters, confirming the per-symbol optimality of classical relay processing in this regime (H2).
    \item A systematic complexity analysis demonstrated that performance \textbf{does not improve monotonically with model size} (H3).
    \item A \textbf{minimal neural relay} (169 parameters) achieved performance comparable to significantly larger models, while intermediate-sized networks exhibited overfitting.
    \item When neural relays were evaluated under equal parameter budgets, architectural differences largely vanished, indicating that \textbf{model capacity}, rather than architectural class, dominates performance for this task (H4).
    \item Sequence-based neural relays such as structured state-space models achieved comparable error performance with improved training efficiency (H5).
    \item From a deployment perspective, a \texttt{hybrid relay} combining neural processing at low SNR with DF at high SNR emerged as the \textbf{most effective strategy}, achieving near-optimal performance with \textbf{minimal overhead}.
    \item On an \textbf{unknown channel} violating classical assumptions (unmodeled intersymbol interference or biased nonlinearity), AF and DF failed to relay reliably --- exhibiting an analytic BER floor of $0.25$, with DF \emph{worsening} as SNR increased --- while the same minimal 170-parameter MLP restored reliable relaying by learning the impairment from data --- including an arbitrary composite cascade of ISI, amplifier nonlinearity, and unknown phase --- tracking a few dB behind a \emph{pilot-aided} Viterbi receiver and, in the fully posterior-free regime (no pilots, no channel prior), matching the best blind classical equalizer while avoiding its instability and online-adaptation cost. Sweeping the intermediate \emph{partial}-posterior case showed the crossover to be governed by channel identifiability and pilot overhead: pilot-aided classical processing wins on long blocks at negligible cost, but the zero-pilot learned relay becomes decisive once blocks are too short or channels too transient to identify from pilots. The learned relay's advantage is specific to \emph{structural} model-class mismatch (memory, nonlinearity, absent pilots) rather than to unknownness per se (H6). Together with the high-SNR result, this delineates precisely where learned relays add value: never above the model-aware optimum, but uniquely robust to \emph{not knowing which model applies} and to the pilot overhead of channel identification. A complexity accounting completed the trade-off: the learned relay's per-symbol cost is constant in modulation order and channel memory and $30$--$90\times$ faster in wall-clock as a vectorized operation, where exact Viterbi MLSE grows as $M^{L}$ and becomes intractable --- so it trades a small BER gap for cost that scales gracefully exactly where classical sequence detection does not.
    \item In the modulation extension, the canonical findings generalized fully to QPSK via I/Q independence, while 16-QAM exposed a structural error floor of per-axis processing; reformulating the relay as a \textbf{joint $N$-class 2D classifier} eliminated the floor and matched DF at high SNR.
\end{enumerate}

Overall, the results indicate that learning-based relays are best used as targeted enhancements rather than universal replacements, with simplicity, parameter efficiency, and hybrid design as key principles for practical cooperative relay systems. All conclusions hold for the \emph{uncoded}, symbol-level canonical setting studied here (Remark~\ref{rem:df-terminology}); extending the comparison to coded block-DF, soft-information (LLR) forwarding relays, other channel families, multi-antenna configurations, and denser constellations (64-QAM and beyond) is identified in Chapter~\ref{sec:discussion-and-conclusions} as the principal direction for future work.

\REV{The summary is faithful to the evidence --- notably, item~8 correctly states the unknown-channel result as bounded (``never above the model-aware optimum'') rather than as a learning-beats-classical headline, which is the intellectually honest version and also the more defensible one. Two suggestions: split item~8 (currently a $\sim$250-word single item that buries the pilot-crossover and complexity results) into three items mirroring the chapter's own structure; and reconcile its ``a few dB behind a pilot-aided Viterbi receiver'' with the precise figures used elsewhere ($\approx$1--1.5~dB on pure ISI, $\approx$2~dB on the composite) --- precision here is free. The same $30$--$90\times$ wall-clock figure repeated here should carry the implementation caveat noted in Section~\ref{sec:complexity-viterbi-mlp}.}
