\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract (English)}
\markboth{Abstract}{Abstract}

\label{sec:abstract-english}


\textbf{Deep Learning for Two-Hop Relay Communication: A Comparative Study of Classical and Neural Network-Based Strategies}

This thesis presents a comparative study of classical and neural network-based relay strategies for two-hop cooperative communication, carried out on a single canonical setup fixed in advance: a SISO link on both hops over i.i.d.\ Rayleigh fast fading, complex baseband, BPSK, and uncoded bit-error rate (BER), with the relay function as the only varied axis. Nine relay methods are evaluated: two classical approaches --- amplify-and-forward (AF) and decode-and-forward (DF) --- and seven neural methods spanning supervised learning (MLP, Hybrid SNR-adaptive relay), generative modeling (VAE, conditional GAN with WGAN-GP), and sequence architectures (Transformer, Mamba S6, Mamba-2 SSD). All results use Monte Carlo simulation (100,000 bits per SNR point, 10 trials, 95\% confidence intervals, Wilcoxon significance testing).

On the canonical setup, the best neural relays outperform AF at low SNR (0--4 dB) but only match symbol-wise DF, which dominates at medium-to-high SNR with zero parameters. A complexity study reveals an inverted-U relationship between model size and performance: a minimal 169-parameter network matches models 100$\times$ larger, while intermediate sizes overfit; at an equal 3,000-parameter budget, architectural differences largely vanish, indicating that capacity, not architecture, drives performance for this task. Mamba-2 SSD trains substantially faster than Mamba S6 at long context with comparable BER.

A scoped modulation study shows the canonical findings transfer fully to QPSK via I/Q independence, while 16-QAM exposes a structural BER floor of per-axis I/Q-split processing that a joint 16-class 2D classifier eliminates, matching DF at high SNR. The principal contribution is a systematic study of learned relaying under \emph{unknown and mismatched} channels. On an unknown 3-tap ISI channel the memoryless classical relays are pinned at an analytic BER floor of 0.25 --- DF worsening as SNR grows --- while the same minimal MLP (here 170 parameters) restores reliable relaying within 1--1.5 dB of genie-CSI Viterbi MLSE with no channel-family knowledge. The result is then bounded from every side: it never beats matched Viterbi; on a composite ISI$\times$nonlinearity$\times$phase cascade it tracks $\sim$2 dB behind pilot-aided MLSE with no impairment model; in the posterior-free regime it matches blind CMA while avoiding the instability of decision-directed MLSE; a pilot-budget sweep locates a sharp identifiability crossover governing when the pilot-free learned relay overtakes pilot-aided classical processing; and a complexity analysis shows the learned relay's per-symbol cost is constant in modulation order and channel memory (where Viterbi grows as $M^{L}$) and $30$--$90\times$ faster in wall-clock. The learned relay thus never beats a matched classical receiver, but occupies a well-defined niche of family-agnostic, identification-free, constant-complexity mitigation.

The recommended deployment is a Hybrid relay combining neural processing at low SNR with DF at high SNR (169 parameters, \textasciitilde0.7 KB, under 3 seconds of training), a joint $N$-class 2D classifier for 16-QAM, and a learned relay wherever the link carries unmodeled memory or bias. The overarching insight is that learned relays never beat the model-aware optimum but are uniquely robust to not knowing which model applies --- and that model capacity must be matched to task structure, not merely task size.

\textbf{Keywords:} Cooperative relay communication, deep learning, two-hop relay, Rayleigh fading, MLP, Mamba, Transformer, VAE, conditional GAN, 16-QAM, intersymbol interference, Viterbi MLSE, blind equalization, bit error rate

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

% \begin{flushright}
% \begin{otherlanguage}{hebrew}
% \textbf{ארכיטקטורות למידה עמוקה לתקשורת ממסר דו-קפיצתית: מחקר השוואתי של אסטרטגיות ממסר קלאסיות ומבוססות רשתות נוירונים}
% גיל צוקרמן
% חיבור זה הוגש כמילוי חלקי של הדרישות לקבלת תואר מגיסטר למדעים (M.Sc.)
% בית הספר להנדסת חשמל — אוניברסיטת תל אביב
% 2026
% \end{otherlanguage}
% \end{flushright}

\clearpage

%% ── Acknowledgments ──────────────────────────────────────────
\cleardoublepage
\phantomsection
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\markboth{Acknowledgments}{Acknowledgments}

I would like to thank my supervisors, Dr. Anatoly Khina and Prof. Raja Giryes at Tel Aviv University, for their guidance and support throughout this research.
This work was conducted in the School of Electrical and Electronic Engineering, Tel Aviv University.


